OneD: increasing reproducibility of Hi-C Samples with abnormal karyotypes

### 2017-06-03 main.tex
Wrote the result section about aberrant karyotypes.

\author[Vidal \textit{et~al}.]{Enrique Vidal\,$^{\text{\sfb 1,2}*}$,
François le Dily\,$^{\text{\sfb 1,2}}$, Javier Quilez\,$^{\text{\sfb
1,2}}$, Ralph Stadhouders\,$^{\text{\sfb 1,2}}$, Yasmina
Cuartero\,$^{\text{\sfb 1,2}}$,Miguel Beato\,$^{\text{\sfb 1,2}}$ and
Guillaume Fillion\,$^{\text{\sfb 1,2}}$}

@@

\begin{table*}
\processtable{Hi-C data sets used in this study.
\label{tab:samples}}
{\begin{tabular}{ccccccc}
  \toprule
  \textbf{Sample ID} & \textbf{Cell type} & \textbf{Application} &
  \textbf{RE} & \textbf{Sequencing core} & \textbf{Set} &
  \textbf{Source} \\
  \midrule
dc3a1e069\_51720e9cf & T47D & \textit{in situ} Hi-C &
  DpnII & CRG & T47D, K562 & NA \\
b1913e6c1\_51720e9cf & T47D & \textit{in situ} Hi-C &
  DpnII & CRG & T47D & NA \\
dc3a1e069\_ec92aa0bb & T47D & \textit{in situ} Hi-C &
  DpnII & CRG & T47D, K562 & NA \\
HindIII\_T0 & T47D & dilution Hi-C &
  HindIII & CRG & T47D & SRR1054341 \\
NcoI\_T0    & T47D & dilution Hi-C &
  NcoI      & CRG & T47D & SRR1054343 \\
ENCLB758KFU & T47D & dilution Hi-C &
  HindIII   & UMass & T47D & ENCLB758KFU \\
ENCLB183QHG & T47D & dilution Hi-C &
  HindIII   & UMass & T47D & ENCLB183QHG \\
HIC069  & K562 & \textit{in situ} Hi-C &
  MboI & Baylor & T47D, K562 &    SRR1658693 \\
HIC070  & K562 & \textit{in situ} Hi-C &
  MboI & Baylor & T47D, K562 &    SRR1658694 \\
HIC071  & K562 & \textit{in situ} Hi-C &
  MboI & Baylor & K562 & SRR1658695,SRR1658696 \\
HIC074  & K562 & \textit{in situ} Hi-C &
  MboI & Baylor & K562 & SRR1658701,SRR1658702 \\
b7fa2d8db\_bfac48760 & B-cell  & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
fc3e8b36a\_7bf1bf374 & ES-cell & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
b7fa2d8db\_7284b867a & B-cell  & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
fc3e8b36a\_38bfd1b33 & ES-cell & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
b7fa2d8db\_58e812fc2 & B-cell  & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
fc3e8b36a\_c990a254e & ES-cell & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
b7fa2d8db\_73f11d923 & B-cell  & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
fc3e8b36a\_4bf044f18 & ES-cell & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
GSM987817 & B-cell  & dilution Hi-C &
  HindIII & UCSC & mm10 & SRR543428-SRR543431 \\
GSM987818 & B-cell  & dilution Hi-C &
  HindIII & UCSC & mm10 & SRR543432-SRR543442 \\
GSM862720 & ES-cell & dilution Hi-C &
  HindIII & UCSC & mm10 & SRR443883-SRR443885 \\
GSM862721 & ES-cell & dilution Hi-C &
  HindIII & UCSC & mm10 & SRR400251-SRR400256 \\
GSM862722 & ES-cell & dilution Hi-C &
  NcoI    & UCSC & mm10 & SRR443886-SRR443888 \\
  \botrule
\end{tabular}}{}
\end{table*}

@@

\subsection{Comparison of Hi-C matrices}

@@

We benchmark \textbf{\textit{oned}}, against methods
\textbf{\textit{vanilla}}, \textbf{\textit{ice}}
\citep{imakaev2012iterative}, \textbf{\textit{caib}}
\citep{wu2016computational} and \textbf{\textit{lgf}}
\citep{hu2012hicnorm, servant2012hitc}. The first three methods correct
biases implicitly, whereas the fourth method does it explicitly.

@@

\subsection{Bias correction}

The principle of OneD is to model explicit Hi-C biases onto a single
variable: the total amount of contacts for each bin of the matrix (see
section~\ref{sec:model}). The reason for this choice of that the total
amount of contacts is approximately proportional to the local copy number.
For instance, a duplicated region in a diploid genome will show on average
a 50\% increase in the number of contacts.

\begin{figure}[!tp]
\centerline{\includegraphics[width=.45\textwidth]{img/figure1.pdf}}
\caption{A. Non linear relationship between the number of restriction
sites per bin and the total number of contacts per bin in T47D. B. Total
number of contacts per bin in chromosome 17 of T47D. Raw (brown) and
corrected (blue) signals are shown. The region that corresponds to the
first 20 Mbp of the chromosome is present in single copy, explaining that
the signal is approximately half compared to the rest of the chromosome.
Also notice that the raw signal is noisier than the corrected one.}
\label{fig:totals}
\end{figure}

Other biases will also affect the total amount of contacts, though not
proportionately. Figure~\ref{fig:totals}A shows the relationship
between the amount of contacts and the availability of retriction enzyme
sites. Four clouds are visible. Each corresponds to a copy number
between one and four. In all of them, the relationship flattens as the
number of restriction sites increases. OneD thus fits a non-linear
relationship between the total amount of contacts and the predictors of
bias (by default the G+C content, the number of restriction sites and the
mappability of the reads).

The biases are estimated for each bin of the genome, and each cell of the
matrix is corrected using equation (\ref{eq:xhat}). The corrected amount
of contacts is still proportional to the copy number, but it is now less
noisy. The correction thus represents a more stable and precise estimation
of the copy number. Figure \ref{fig:totals}B shows the corrected number of
contacts along chromosme 17 of a T47D sample. The corrected contacts have
a narrower range than the raw contacts.

We tested the validity of this approach against the Catalogue Of Somatic
Mutations In Cancer \citep[COSMIC,][]{forbes2010cosmic}. Figure
\ref{fig:copy_number} shows the Pearson correlation between the corrected
number of contacts and the copy number estimation for both T47D and K562
cell lines. OneD achieves the highest correlations in both cases.
Actually, it is the only method that increases the agreement with the
external data. The other methods, including caICB, only make the
normalized data more divergent than the raw counts.  Similar results were
obtained using Spearman correlation (Supplementary Figure 2).

Taken together, these results indicate that OneD is an efficient way to
estimate the local copy number in order to correct for it.

@@

\subsection{Samples with aberrant karyotypes}

Focusing on the Pearson correlation of observed over expected counts,
methods based on the iterative correction (mainly \textit{ice} and
\textit{caib} but also \textit{vanilla}) present the undesirable feature
of higher correlations between unrelated samples than using raw data (top
left panels of Figures \ref{fig:aberrant}A and \ref{fig:aberrant}C,
Supplementary Figure 3). Among the comparisons of cells with the same
type, \textit{oned} is the only method outperforming uncorrected data when
comparing experiments using different protocols (top right panel of Figure
\ref{fig:aberrant}A) and it is among the top performant when comparing
experiments using the same protocol (rest of the panels in Figures
\ref{fig:aberrant}A and \ref{fig:aberrant}C). Consequently, our proposal
is the one that better discriminates unrelated samples form similar ones
(as it can be seen in the ROC curves and AUCs of Figures
\ref{fig:aberrant}B and D).

**
In an attempt to further summarize the results, we reasoned that the
correlation between unrelated samples (\textit{i.e} different cell types)
should be lower than the correlation between related samples (\textit{i.e}
same cell types), so we could re-interpret each correlation metric as a
classifier score.
**

Complementary analysis using other metrics (Supplementary Figures 3, 4 and
5) yielded similar results. Notably, Spearman correlation of contacts
presents the worst performance for all scenarios and seems to be a poor
choice as a metric to compare HiC matrices.

\begin{figure}
	\centerline{\includegraphics[width=.50\textwidth]{img/correlation_aberrant_figure3.pdf}}
	\caption{
    Results of the comparison between samples with aberrant karyotype.
Left panels: Average difference and 95\% CI between each correction method
(Y axis) and the raw data (brown dashed line) for T47D (A) and K562 (C)
sets. Right panels: Corresponding ROC curves and AUCs for T47D (B) and
K562 (D) sets. All results in this figure are based on Pearson
correlations between the observed over expected counts. Our method
increases reproducibility of similar samples while being able to
discriminate different ones.  }\label{fig:aberrant}
\end{figure}

\subsection{Samples with normal karyotypes}

>>>>

\author[Vidal \textit{et~al}.]{Enrique Vidal\,$^{\text{\sfb 1,2}*}$,
François le Dily\,$^{\text{\sfb 1,2}}$, Javier Quilez\,$^{\text{\sfb
1,2}}$, Ralph Stadhouders\,$^{\text{\sfb 1,2}}$, Yasmina
Cuartero\,$^{\text{\sfb 1,2}}$, Miguel Beato\,$^{\text{\sfb 1,2}}$ and
Guillaume Fillion\,$^{\text{\sfb 1,2}}$}

@@

\begin{table*}
\processtable{Hi-C data sets used in this study.
\label{tab:samples}}
{\begin{tabular}{ccccccc}
  \toprule
  \textbf{Sample ID} & \textbf{Cell type} & \textbf{Application} &
  \textbf{RE} & \textbf{Sequencing core} & \textbf{Set(s)} &
  \textbf{Source} \\
  \midrule
dc3a1e069\_51720e9cf & T47D & \textit{in situ} Hi-C &
  DpnII & CRG & T47D, K562 & NA \\
b1913e6c1\_51720e9cf & T47D & \textit{in situ} Hi-C &
  DpnII & CRG & T47D & NA \\
dc3a1e069\_ec92aa0bb & T47D & \textit{in situ} Hi-C &
  DpnII & CRG & T47D, K562 & NA \\
HindIII\_T0 & T47D & dilution Hi-C &
  HindIII & CRG & T47D & SRR1054341 \\
NcoI\_T0    & T47D & dilution Hi-C &
  NcoI      & CRG & T47D & SRR1054343 \\
ENCLB758KFU & T47D & dilution Hi-C &
  HindIII   & UMass & T47D & ENCLB758KFU \\
ENCLB183QHG & T47D & dilution Hi-C &
  HindIII   & UMass & T47D & ENCLB183QHG \\
HIC069  & K562 & \textit{in situ} Hi-C &
  MboI & Baylor & T47D, K562 &    SRR1658693 \\
HIC070  & K562 & \textit{in situ} Hi-C &
  MboI & Baylor & T47D, K562 &    SRR1658694 \\
HIC071  & K562 & \textit{in situ} Hi-C &
  MboI & Baylor & K562 & SRR1658695,SRR1658696 \\
HIC074  & K562 & \textit{in situ} Hi-C &
  MboI & Baylor & K562 & SRR1658701,SRR1658702 \\
b7fa2d8db\_bfac48760 & B-cell  & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
fc3e8b36a\_7bf1bf374 & ES-cell & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
b7fa2d8db\_7284b867a & B-cell  & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
fc3e8b36a\_38bfd1b33 & ES-cell & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
b7fa2d8db\_58e812fc2 & B-cell  & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
fc3e8b36a\_c990a254e & ES-cell & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
b7fa2d8db\_73f11d923 & B-cell  & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
fc3e8b36a\_4bf044f18 & ES-cell & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
GSM987817 & B-cell  & dilution Hi-C &
  HindIII & UCSC & mm10 & SRR543428-SRR543431 \\
GSM987818 & B-cell  & dilution Hi-C &
  HindIII & UCSC & mm10 & SRR543432-SRR543442 \\
GSM862720 & ES-cell & dilution Hi-C &
  HindIII & UCSC & mm10 & SRR443883-SRR443885 \\
GSM862721 & ES-cell & dilution Hi-C &
  HindIII & UCSC & mm10 & SRR400251-SRR400256 \\
GSM862722 & ES-cell & dilution Hi-C &
  NcoI    & UCSC & mm10 & SRR443886-SRR443888 \\
  \botrule
\end{tabular}}{}
\end{table*}

@@

\subsection{Comparison of Hi-C matrices}
\label{sec:comp}

@@

We benchmarked \textbf{\textit{oned}}, against \textbf{\textit{vanilla}},
\textbf{\textit{ice}} \citep{imakaev2012iterative}, \textbf{\textit{caib}}
\citep{wu2016computational} and \textbf{\textit{lgf}}
\citep{hu2012hicnorm, servant2012hitc}. The first three methods correct
biases implicitly, whereas the fourth method does it explicitly.

@@

\subsection{Correction method}

The principle of OneD is to explicitly model Hi-C biases on a single
variable: the total amount of contacts for each bin of the matrix (see
\ref{sec:model}). The reason for this choice of that the total amount of
contacts is approximately proportional to the local copy number. For
instance, a duplicated region in a diploid genome will show on average a
50\% increase in the number of contacts. Discontinuities of the amount of
contacts thus correspond to changes of the copy number.

\begin{figure}[!tp]
\centerline{\includegraphics[width=.45\textwidth]{img/figure1.pdf}}
\caption{A. Non linear relationship between the number of restriction
sites per bin and the total number of contacts per bin in T47D. B. Total
number of contacts per bin in chromosome 17 of T47D. Raw (brown) and
corrected (blue) signals are shown. The region that corresponds to the
first 20 Mbp of the chromosome is present in single copy in this clone of
T47D, explaining that the signal is approximately half compared to the
rest of the chromosome.  Also notice that the raw signal is noisier than
the corrected one.}
\label{fig:totals}
\end{figure}

Other biases affect the total amount of contacts in a countinuous but not
necessarily linear way. Figure~\ref{fig:totals}A shows the relationship
between the amount of contacts and the number of retriction enzyme sites
in T47D, a breast cancer cell line with aberrant and unstable karyotype.
Four clouds are visible. Each corresponds to a copy number between one and
four. In all of them, the relationship flattens as the number of
restriction sites increases. For this reason, OneD fits a non-linear
relationship between the total amount of contacts and the known sources of
bias (by default the G+C content, the number of restriction sites and the
mappability of the reads).

The biases are estimated genome-wide and each cell of the matrix is then
corrected using equation (\ref{eq:xhat}). Note that the corrected amount
of contacts is still proportional to the copy number. The correction thus
represents a more stable and precise estimation of the copy number.
Figure~\ref{fig:totals}B shows the corrected number of contacts along
chromosme 17 of a T47D. OneD greatly reduces the wiggling of the total
amount of contacts.

We tested the validity of this approach against the Catalogue Of Somatic
Mutations In Cancer \citep[COSMIC,][]{forbes2010cosmic}. Figure
\ref{fig:copy_number} shows the Pearson correlation between the corrected
number of contacts and the copy number estimation for both T47D and K562
cell lines. Similar results were obtained using Spearman correlation
(Supplementary Figure 2). All the methods except OneD decrease the
agreement with the copy number because they partially correct this bias.
In contrast, OneD enhances the conformity of the signal with the copy
number. Not correcting for variable copy number at that stage may seem to
defeat the purpose of a normalization method, but we will see below how
this can lead to better performance.

@@

\subsection{Aberrant karyotypes}

We first benchmarked the performance of OneD on biological samples with an
aberrant karyotype. A good normalization method should increase the
similarity between biological replicates by reducing irrelevant
experimental variation, such as batch effects, laboratory of origin and
protocol variations. Similarly, a good normalization should decrease the
similarity between samples that received a different treatment in order to
enhance the biological differences.

We assembled two Hi-C data sets obtained from T47D and K562, two cancer
cell lines with aberrant karyotypes. In each set we spiked two
samples from the other cell line (see Table~\ref{tab:samples}) in order to
introduce biological variability. We measured the Pearson correlation of
observed over expected counts to compare matrices before and after
normalization by different methods (see \ref{sec:comp}). This gave an
indication of the impact of a given normalization method. The results are
summarized in Figure~\ref{fig:aberrant}.

The \textit{caib} and \textit{ice} methods increased the similarity
between the different cell lines (Figures~\ref{fig:aberrant}A and
\ref{fig:aberrant}C and Supplementary Figure~3). This is an undesirable
effect, as it obscures the biological variability. In the same vein, these
methods decreased the similarity between samples that received the same
treatment (Figures~\ref{fig:aberrant}A), suggesting that the normalization
process is detrimental to the biological signal in these two cases. The
method \textit{vanilla} followed the same trend but to a lesser extent,
consistent with the fact that it consists of a single \textit{ice}
iteration.

OneD was the only method to increase the similarity between experiments
carried out on the same material but with a different protocol
(Figure~\ref{fig:aberrant}A). In these conditions, \textit{lgf} was
decreasing the similarity. Taken together, the results show that OneD
enhanced biological variation while decreasing experimental variation.

An important application of normalization methods in experimental setups
is to identify outliers. We thus investigated the capacity of the
different methods to help identify the samples from the other cell type
spiked in the data set. We interpreted the pairwise comparison scores as
classifier scores and summarized the results with a ROC curve
(Figures~\ref{fig:aberrant}B and D).

All the methods, including the absence of normalizaton, succeeded in
recognizing the T47D outliers among the K562 samples, but recognizing the
K562 outliers among the T47D samples proved more challenging. OneD
increased the discrimination power compared to the raw matrices, but all
the other methods decreased it. Actually, they performed little better
than random on this task.

Using the other metrics described in \ref{sec:comp} yielded similar
results (Supplementary Figures 3, 4 and 5). Note that Spearman correlation
of contacts presented the worst performance for all scenarios, and it thus
seems to be a poor choice as a metric to compare HiC matrices.

Taken together, these results show that OneD enhances biological variation
and reduces experimental noise on samples with an aberrant karyotype.

\begin{figure}
\centerline{\includegraphics[width=.50\textwidth]{img/correlation_aberrant_figure3.pdf}}
\caption{
Results of the comparison between samples with aberrant karyotype.  A and
B. Average changes compared to raw on the T47D and K562 sets.  The bars
represent 95\% confidence intervals centered on the mean difference of the
correlation score between a given correction method and the raw data. The
brown dashed line indicates the value of the average score on raw matrices
(set to 0). C and D. ROC curves on the T47D and K562 sets. The areas under
the curve are indicated in the bottom right corner. The color code is the
same as panels A and B. The brown lines correspond to raw matrices.  All
results in this figure are based on Pearson correlations between the
observed over expected counts.  Our method increases reproducibility of
similar samples while being able to discriminate different ones.}
\label{fig:aberrant}
\end{figure}

\subsection{Normal karyotypes}
